\bf Manifold Clustering in the Generalized Random Dot Product Graph

### 2022-05-05 draft/draft.Rmd
changes based on meeting earlier today

\begin{example}[Hierarchical $\ER$ model]
In the $\ER$ model, the edge probability matrix has a fixed value $[P_{ij}] \equiv p \in [0, 1]$. 
Suppose instead that each $P_{ij}$ is drawn from a distribution with density $f(p) = \frac{1}{\frac{\pi}{2} - 1} \Big(\frac{1}{\sqrt{1 - p^2}} - 1 \Big)$.

Then it can be shown that this model is equivalent to a random dot product graph with a two dimensional latent space in which latent vectors are sampled uniformly along the curve $g(t) = \begin{bmatrix} \cos(\frac{\pi}{2} t) & \sin(\frac{\pi}{2} t) \end{bmatrix}^\top$, $0 \leq t \leq 1$. 
\end{example}

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \GRDPG_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition}

In practice, we often define each manifold $\mathcal{M}_k$ by a continuous function $g_k : [0, 1]^r \to \mathcal{X}$, where $1 \leq r < d$ is the dimensionality of manifold $\mathcal{M}_k$. 
To complete the probability model, we define a probability distribution $F$ with support $[0, 1]^r$ and draw the community memberships from a multinomial distribution.

The full mixture model can be described as follows:

1. Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$. 
2. Draw latent vectors by first drawing each $t_1, ..., t_n \iid F$ and then computing each $x_i = g_{z_i}(t_i)$. 
3. Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$, 
and draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$.

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two one-dimensional manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Fig \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.

@@

@@

\begin{proof}
Let $H^{(k)}$ be the event that lemma \ref{lem:one-hypercube} holds for $\mathcal{C}_k$. Then $H^{(k)} = H_{n_k}$ where $H_n$ is defined as in \ref{lem:one-hypercube}. 
Then $P(H'_n) = P(H_{n_1} \text{ and } ... \text{ and } H_{n_K})$.

>>>>

\begin{example}[Hierarchical $\ER$ model]
In the $\ER$ model, the edge probability matrix has a fixed value $[P_{ij}] \equiv p \in [0, 1]$.

Suppose that we have a random dot product graph in which the latent space is $\mathbb{R}^2$ and latent vectors are drawn uniformly from the quarter circle defined by $g(t) = \begin{bmatrix} \cos(\frac{\pi}{2} t) & \sin(\frac{\pi}{2} t) \end{bmatrix}^\top$, $0 \leq t \leq 1$. 
Then it can be shown that in this model, instead of a fixed $P_{ij} = p$, the edge probabilities are distributed with density $f(p) = \frac{2}{\pi - 2} \Big(\frac{1}{\sqrt{1 - p^2}} - 1 \Big)$. 
\end{example}

By changing the latent structure from a point mass to a curve, we are able to come up with more flexible Bernoulli graph models in which edge probabilities follow more general probability distributions. 
Community structure then can be added by sampling latent vectors from multiple curves. 
Then the adjacency spectral embedding of the resulting graph allows us to recover that community structure. 
This is illustrated in the following example.

\begin{example}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \iid \Uniform(0, 1)$ and $z_1, ..., z_n \iid \Multinomial(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \RDPG(X)$. Fig. \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

We now formally define the manifold block model.

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n \geq 1$ be integers.
Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$ each by continuous function $g_k : [0, 1] \to \mathcal{X}$, 
and probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
Then the following mixture model is a manifold block model:

\begin{enumerate}
  \item Draw labels $z_1, ..., z_n \iid \Multinomial(\alpha_1, ..., \alpha_K)$. 
  \item Draw latent vectors by first drawing each $t_i \indep F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
  \item Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$, and draw $A \sim \RDPG(X)$ or $A \sim \GRDPG_{p,q}(X)$. 
\end{enumerate}
\end{definition}

@@

\begin{proof}
Let $H^{(k)}$ be the event that lemma \ref{lem:one-hypercube} holds for $\mathcal{C}_k$. Then $H^{(k)} = H_{n_k}$ where $H_n$ is defined as in lemma \ref{lem:one-hypercube}. 
Then $P(H'_n) = P(H_{n_1} \text{ and } ... \text{ and } H_{n_K})$.

### 2022-05-05 draft/draft.tex
changes based on meeting earlier today

\begin{example}[Hierarchical $\text{Erd\"{o}s-R\'{e}nyi}$ model]
In the $\text{Erd\"{o}s-R\'{e}nyi}$ model, the edge probability matrix has a fixed value $[P_{ij}] \equiv p \in [0, 1]$. 
Suppose instead that each $P_{ij}$ is drawn from a distribution with density $f(p) = \frac{1}{\frac{\pi}{2} - 1} \Big(\frac{1}{\sqrt{1 - p^2}} - 1 \Big)$.

Then it can be shown that this model is equivalent to a random dot product graph with a two dimensional latent space in which latent vectors are sampled uniformly along the curve $g(t) = \begin{bmatrix} \cos(\frac{\pi}{2} t) & \sin(\frac{\pi}{2} t) \end{bmatrix}^\top$, $0 \leq t \leq 1$. 
\end{example}

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $K \geq 1$, and $n \geq 1$ be integers.
If there are manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$, and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ such that each $x_i \in \mathcal{M}_k$ for some $k \in [K]$, then $A \sim \mathrm{GRDPG}_{p,q}(X; \rho_n)$ is a \emph{manifold block model}.
\end{definition}

In practice, we often define each manifold \(\mathcal{M}_k\) by a
continuous function \(g_k : [0, 1]^r \to \mathcal{X}\), where
\(1 \leq r < d\) is the dimensionality of manifold \(\mathcal{M}_k\). To
complete the probability model, we define a probability distribution
\(F\) with support \([0, 1]^r\) and draw the community memberships from
a multinomial distribution.

The full mixture model can be described as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw labels
  \(z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)\).
\item
  Draw latent vectors by first drawing each
  \(t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}F\) and then computing
  each \(x_i = g_{z_i}(t_i)\).
\item
  Let \(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\), and
  draw \(A \sim \mathrm{RDPG}(X)\) or
  \(A \sim \mathrm{GRDPG}_{p,q}(X)\).
\end{enumerate}

\begin{example}
Let $p = 2$, $q = 0$, $K = 2$, and $r = 1$. 
Define two one-dimensional manifolds by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$ and $z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \mathrm{RDPG}(X)$. Fig \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph.

@@

>>>>

\begin{example}[Hierarchical $\text{Erd\"{o}s-R\'{e}nyi}$ model]
In the $\text{Erd\"{o}s-R\'{e}nyi}$ model, the edge probability matrix has a fixed value $[P_{ij}] \equiv p \in [0, 1]$.

Suppose that we have a random dot product graph in which the latent space is $\mathbb{R}^2$ and latent vectors are drawn uniformly from the quarter circle defined by $g(t) = \begin{bmatrix} \cos(\frac{\pi}{2} t) & \sin(\frac{\pi}{2} t) \end{bmatrix}^\top$, $0 \leq t \leq 1$. 
Then it can be shown that in this model, instead of a fixed $P_{ij} = p$, the edge probabilities are distributed with density $f(p) = \frac{2}{\pi - 2} \Big(\frac{1}{\sqrt{1 - p^2}} - 1 \Big)$. 
\end{example}

By changing the latent structure from a point mass to a curve, we are
able to come up with more flexible Bernoulli graph models in which edge
probabilities follow more general probability distributions. Community
structure then can be added by sampling latent vectors from multiple
curves. Then the adjacency spectral embedding of the resulting graph
allows us to recover that community structure. This is illustrated in
the following example.

\begin{example}
Define two one-dimensional manifolds in $\mathbb{R}^2$ by $f_1(t) = \begin{bmatrix} \cos(\frac{\pi}{3} t) & \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$ and $f_2(t) = \begin{bmatrix} 1 - \cos(\frac{\pi}{3} t) & 1 - \sin(\frac{\pi}{3} t) \end{bmatrix}^\top$.
Draw $t_1, ..., t_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Uniform}(0, 1)$ and $z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\frac{1}{2}, \frac{1}{2})$, and compute latent vectors $x_i = f_{z_i}(t_i)$, which are collected in data matrix $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$. 
Finally, let $A \sim \mathrm{RDPG}(X)$. Fig. \ref{fig:example1} shows the latent configuration drawn from this latent distribution, a random dot product graph drawn from the latent configuration, and the adjacency spectral embedding of the graph. 
Although the community structure is not obvious from the graph, the embedding shows a clear separation between the two communities.

@@

We now formally define the manifold block model.

\begin{definition}[Manifold block model]
Let $p, q \geq 0$, $d = p + q \geq 1$, $1 \leq r < d$, $K \geq 2$, and $n \geq 1$ be integers.
Define manifolds $\mathcal{M}_1, ..., \mathcal{M}_K \subset \mathcal{X}$ for $\mathcal{X} = \{x, y \in \mathbb{R}^d : x^\top I_{p,q} y \in [0, 1] \}$ each by continuous function $g_k : [0, 1] \to \mathcal{X}$, 
and probability distributions $F_1, ..., F_K$ each with support $[0, 1]^r$. 
Then the following mixture model is a manifold block model:

\begin{enumerate}
  \item Draw labels $z_1, ..., z_n \stackrel{\mathrm{iid}}{\sim}\mathrm{Multinomial}(\alpha_1, ..., \alpha_K)$. 
  \item Draw latent vectors by first drawing each $t_i \stackrel{\mathrm{ind}}{\sim}F_{z_i}$ and then computing each $x_i = g_{z_i}(t_i)$. 
  \item Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$, and draw $A \sim \mathrm{RDPG}(X)$ or $A \sim \mathrm{GRDPG}_{p,q}(X)$. 
\end{enumerate}
\end{definition}
